---
title: "Integrating biological status, habitat status, and climate change to inform resilience planning for Pacific salmon"
author: "Stephanie Peacock, et al."
date: '`r Sys.Date()`'
output: 
  html_document:
    toc: true
    toc_float: true
    fig_width: 8
    fig_height: 7
    fig_caption: true
bibliography: bib.json
csl: canadian-journal-of-fisheries-and-aquatic-sciences.csl 
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(plotly)
library(viridisLite)

# Functions to connect to PSF database
source("https://raw.githubusercontent.com/salmonwatersheds/population-indicators/refs/heads/master/code/functions_general.R")

# PSf colours and species colours
source("https://raw.githubusercontent.com/salmonwatersheds/population-indicators/refs/heads/master/code/colours.R")
names(status_cols)[1:3] <- c("good", "fair", "poor")

```

# Background

* Pacific salmon are experiencing widespread declines throughout western North America.
* The current management paradigm is focused on rebuilding stocks that have reached critically low abundances by identifying and attempting to address factors that led to their decline or may be inhibiting their recovery.
* Although considerations of habitat degradation and climate climate change exposure may factor into recovery plans, there is a need for a proactive approach to salmon management that considers all salmon populations, their habitats, and the increasing threat of climate changes throughout their lifecycles.

# Data compilation
* PSF has assessed the status of Pacific salmon Conservation Units and their freshwater habitats throughout B.C. and the Yukon [@ConnorsEtAl2026DemocratizingDataSupport].
* More recently, climate change exposure of 60 CUs in the Fraser has been quantified [@PeacockEtAl2026QuantifyingExposurePacific].
* We transformed biological and habitat status into numeric scores between 1 (good biological status or low threat of habitat degradation) and 3 (poor biological status or high threat of degradation):
  * Biological status scores were calculated from the probabilities (accounting for benchmark uncertainty) of good, fair, and poor status outcomes as: $B = (1 \times \text{prob(good)}) + (2 \times \text{prob(fair)}) + (3 \times \text{prob(poor)})$
  * Habitat status scores were calculated from the proportion of spawning habitat area at low, moderate or high risk of degradation based on cumulative pressures from seven habitat pressure indicators: $H = (1 \times \text{prop(low threat)}) + (2 \times \text{prop(moderate threat)}) + (3 \times \text{prop(high threat)})$
* Climate change exposure was a score between 0 and 1 indicating the proportion of time, averaged among life stages, that a CU is exposed to climate variables above or below optimal thresholds based on mid-century (2040-2069) projected climate under a moderate emissions scenario (RCP 4.5; see @PeacockEtAl2026QuantifyingExposurePacific).
* All three scores (biological status, habitat status, and climate change exposure) were scaled to a mean of zero and standard deviation of 1 for subsequent analyses.


```{r}
#------------------------------------------------------------------------------
# Read in data (see code/data-processing.R for compilation steps)
#------------------------------------------------------------------------------

data <- read.csv("output/CU_pophabclim_scores.csv")

# Scale data for plotting and cluster analysis
scaled_data <- data %>%
  mutate(pop_score = scale(pop_score),
         hab_score = scale(hab_score),
         clim_score = scale(clim_score))

```

# Data visualization 

* Biological status, habitat status, and climate change exposure for each of 49 CUs can be visualized on a 3D plot.

```{r, echo=FALSE, caption = "Plot of biological status, habitat status, and climate change exposure for 49 CUs in the Fraser River basin. CU names appear on hover, and points are coloured by species (see legend)."}
#------------------------------------------------------------------------------
# 3D plot by species
#------------------------------------------------------------------------------

scaled_data$species_name <- factor(scaled_data$species_name, levels = names(species_cols_dark))
scaled_data <- scaled_data %>% arrange(species_name, pop_score)

# By species
fig <- plot_ly(
  scaled_data, # Need to arrange to avoid misrep
  x = ~pop_score, y = ~hab_score, z = ~clim_score, 
  color = ~species_name, colors = species_cols_light, 
  text = ~paste(cu_name_pse, species_name), hoverinfo = 'text') %>%
  layout(scene = list(
    xaxis = list(title = "Biological"),
    yaxis = list(title = "Habitat"),
    zaxis = list(title = "Climate change",
                 aspectmode = 'data') # Avoids re-scaling
  )) %>%
  add_markers() 
fig # Display the plot
```
* The current "recovery paradigm" is focused on identifying threats and recovery strategies for CUs that have poor biological status (identified in red below). For these CUs, habitat and climate change considerations are often looked at in the context of factors that may have contributed to declines or may be limiting recovery.

```{r, echo=FALSE, caption = "Plot of biological status, habitat status, and climate change exposure for 49 CUs in the Fraser River basin, highlighting CUs in red that have poor biological status and may be targets for recovery planning under the current paradigm."}
#------------------------------------------------------------------------------
# 3D plot by biological status
#------------------------------------------------------------------------------
scaled_data$biostatus <- factor(scaled_data$biostatus, levels = c("good", "fair", "poor"))
scaled_data <- scaled_data %>% arrange(biostatus, pop_score)

fig <- plot_ly(
  scaled_data, 
  x = ~pop_score, y = ~hab_score, z = ~clim_score, 
  color = ~biostatus, colors = status_cols, 
  text = ~paste(cu_name_pse, species_name), hoverinfo = 'text')
fig <- fig %>% 
  layout(scene = list(
  xaxis = list(title = "Biological"),
  yaxis = list(title = "Habitat"),
  zaxis = list(title = "Climate change")
)) %>%
  add_markers() 
fig # Display the plot
```

* A more forward-looking approach may be to consider "resilience strategies" for groups of CUs depending on where they lie in the 3D space of biological status, habitat status, and climate change exposure. 
  * This approach considers the potential threats *and opportuntities* to promote positive outcomes for salmon in the face of climate change.
  * The potential trade-offs are made explicit; for example between recovering CUs with poor status and high climate change exposure and protecting those with fair or good status and low climate change exposure.
  
## Identifying CU clusters
* We can identify strategies associated with different regions of the 3D space, although this may ignore the actual distribution of CUs in that space (i.e., there may be a need to consider multiple strategies in regions with a lot of CUs, whereas there may be few or no CUs that exist in other areas).
* Alternatively, we can look at the actual data and identify natural clusters of CUs in the 3D space.
* K-means clustering can identify groups of CUs that are closest to each other in terms of biological status, habitat status, and climate change exposure, and then we can consider strategies that would be best-suited to advancing resilience of each cluster.
* Initially, we can set the target number of clusters (say, 5), but we may want to extend this to determine the optimal number of clusters. (Note that there may be other constraints on the desired number of clusters; it is not necessarily helpful to identify unique strategies for each CU, nor do we want just one cluster to emerge.)

```{r, echo = TRUE}
#-------------------------
# K-means cluster analysis
#-------------------------

# Set target number of clusters
n_clust <- 5

# Run K-means with n_clust
set.seed(123) # for reproducibility
kmeans_result <- kmeans(
  scaled_data %>% select(pop_score, hab_score, clim_score), 
  centers = n_clust, 
  nstart = 25)

# Add cluster assignments back to original data
scaled_data$k_cluster <- as.factor(kmeans_result$cluster)

# View results
print(kmeans_result$centers) # Interpretation of cluster characteristics

```

* We can visualize the five clusters in the original 3D plot:
```{r, echo = FALSE, warning=FALSE, caption = "Plot of biological status, habitat status, and climate change exposure for 49 CUs in the Fraser River basin, colouring CU points according to how they cluster in 3D space."}
#------------------------------------------------------------------------------
# 3D plot by biological status
#------------------------------------------------------------------------------

scaled_data <- scaled_data %>% arrange(k_cluster, pop_score)

fig <- plot_ly(
  scaled_data, 
  x = ~pop_score, y = ~hab_score, z = ~clim_score, 
  color = ~k_cluster, colors = viridis(n = n_clust),
  text = ~paste(cu_name_pse, species_name), hoverinfo = 'text')
fig <- fig %>% 
  layout(scene = list(
  xaxis = list(title = "Biological"),
  yaxis = list(title = "Habitat"),
  zaxis = list(title = "Climate change")
)) %>%
  add_markers() 
fig # Display the plot
```

### Alternative cluster approach
* More detail on the differentiation among CUs can be gleaned from a hierarchical cluster analysis:

```{r, echo = TRUE, fig.width=10, fig.height=10}
#-------------------------------
# Hierarchical cluster analysis
#-------------------------------

# Compute distance matrix
d <- dist(scaled_data %>% select(pop_score, hab_score, clim_score), method = "euclidean")

# Perform hierarchical clustering ("ward.D2" = Ward's method is usually best)
h_clust <- hclust(d, method = "complete")

# Plot the dendrogram
plot(h_clust, labels = paste(scaled_data$cu_name_pse, scaled_data$species_name), main = "Dendrogram")

# Cut the tree into a specific number of groups (e.g., n_clust to correspond to k-means)
rect.hclust(h_clust, k = n_clust, border = viridis(n = n_clust)) # Add boxes to plot

# Add tree cut clusters to data frame
scaled_data$h_cluster <- as.factor(cutree(h_clust, k = n_clust))
```

  * This hierarchical clustering produces slightly different clusters than k-means, but they should be the same provided the same target is used. Something to look into...
```{r , echo=FALSE, caption = "Plot of biological status, habitat status, and climate change exposure for 49 CUs in the Fraser River basin, colouring CU points according to how they cluster in 3D space."}
#------------------------------------------------------------------------------
# 3D plot by biological status
#------------------------------------------------------------------------------

scaled_data <- scaled_data %>% arrange(h_cluster, pop_score)

fig <- plot_ly(
  scaled_data, 
  x = ~pop_score, y = ~hab_score, z = ~clim_score, 
  color = ~h_cluster, colors = viridis(n = n_clust),
  text = ~paste(cu_name_pse, species_name), hoverinfo = 'text')
fig <- fig %>% 
  layout(scene = list(
  xaxis = list(title = "Biological"),
  yaxis = list(title = "Habitat"),
  zaxis = list(title = "Climate change")
)) %>%
  add_markers() 
fig # Display the plot

scaled_data <- scaled_data %>% arrange(k_cluster, pop_score) # Arrange data back as foir k-means

```

## Example CUs
* What are three example CUs for each cluster? Choose the 3 CUs that have the smallest distance to the centre as examples.

```{r, include = TRUE, echo = TRUE}

# Calculate distance for each CU from it's k-means cluster centre
scaled_data$dist_from_centre <- NA

for(j in 1:n_clust){
  
  scaled_data.j <- scaled_data %>% filter(k_cluster == j)
  
  n.j <- nrow(scaled_data.j)
  
  mat.j <- rbind(kmeans_result$centers[j,],
               scaled_data.j %>% select(pop_score, hab_score, clim_score))
  
  d.j <- dist(mat.j, method = "euclidean")[1:n.j]
  
  scaled_data$dist_from_centre[which(scaled_data$k_cluster == j)] <- d.j
}

# Extract 3 CUs closest to centre
example_CUs <- data.frame(
  cluster = rep(c(1:n_clust), each = 3),
  species_name = NA,
  cuid = NA,
  cu_name_pse = NA
)

for(j in 1:n_clust){
  scaled_data.j <- scaled_data %>% filter(k_cluster == j) %>% arrange(dist_from_centre)
  
  example_CUs[(j-1)*3 + c(1:3), 1] <- j
  example_CUs[(j-1)*3 + c(1:3), 2] <- as.character(scaled_data.j$species_name[1:3])
  example_CUs[(j-1)*3 + c(1:3), 3] <- scaled_data.j$cuid[1:3]
  example_CUs[(j-1)*3 + c(1:3), 4] <- scaled_data.j$cu_name_pse[1:3]
  
}

print(example_CUs)
  
```

# Resilience strategies for clusters

* For each of the five clusters identified by k-means, we can think about what kinds of strategies may be best suited to promoting resilience.
  * (Remember: low values of the scores are good biological status, low risk habitats, and low climate exposure.)

```{r table_cluster, include = TRUE, echo = FALSE}
options(width = 400)
library(reactable)
library(htmltools)
library(fontawesome)

colRamp <- colorRampPalette(status_cols[1:3])(n = 100)

strat <- data.frame(kmeans_result$centers) %>%
  mutate(cluster_no = c(1:5), 
         cluster_size = table(scaled_data$k_cluster),
         pop_score = round(pop_score, 3),
         hab_score = round(hab_score, 3),
         clim_score = round(clim_score, 3)) %>%
  select(cluster_no, pop_score, hab_score, clim_score, cluster_size) %>%
  mutate(summary = c(
    "Fair to poor biological status and moderate risk of habitat degradation.", 
    "Good biological status and low climate exposure, but high risk of habitat degradation.", 
    "Poor biological status and high climate change exposure.", 
    "High risk of habitat degradation, with fair biological status and above-average climate exposure.",
    "Relatively high climate exposure but intact habitats may buffer against change; good biological status." 
     
    ),
         strategy = c(
           "Efforts in both habitat and population recovery are worthwhile as climate exposure is relatively low.", 
           "Habitat restoration may solidify these as salmon strongholds.", 
           "Despite moderate risk to habitats, recovering these populations will require significant ongoing investment due to worsening impacts of climate change" ,
           "Habitat restoration is needed, but the return on investments may be less than for Cluster 2 owing to low current abundance and higher climate change exposure.",
           "Protect habitat and take actions (e.g., reduce fishing) to improve biological status." 
           ))

rownames(strat) <- NULL

reactable::reactable(
      strat,
      columns = list(
        cluster_no = colDef(
          maxWidth = 50,
          style = function(value){
            color <- c("white", "black")[c(1,1,1,2,3)][value]
            background <- viridis(n = n_clust)[value]
            list(color = color, background = background, font = "bold")
          }
        ),
        pop_score = colDef(
          maxWidth = 60,
          style = function(value){
            color <- colRamp[findInterval(value, seq(-2, 2, length.out = 100))]
            list(background = color)
          }
          
        ),
        hab_score = colDef(
          maxWidth = 60,
          style = function(value){
            color <- colRamp[findInterval(value, seq(-2, 2, length.out = 100))]
            list(background = color)
          }
          
        ),
        clim_score = colDef(
          maxWidth = 60,
          style = function(value){
            color <- colRamp[findInterval(value, seq(-2, 2, length.out = 100))]
            list(background = color)
          }
          
        ),
        cluster_size = colDef(maxWidth = 60),
          summary = colDef(maxWidth = 200),
        strategy = colDef(maxWidth = 200)
      ),
      bordered = TRUE,
      highlight = TRUE,
      striped = TRUE,
      resizable = TRUE, 
      wrap = TRUE,
      searchable = TRUE,
      style = list(fontSize = "1.25rem"),
      showSortIcon = TRUE
)
```

# Supplemental output
Below is downloadable table of the biological status scores, habitat status scores, and climate change exposure scores for each CU along with the cluster assigned by k-means (`k_cluster`) or by the hierarchical clustering (`h_cluster`).

```{r table, include = TRUE, echo = FALSE}
options(width = 400)


tab <- scaled_data %>% select(
  species_name, cu_name_pse, biostatus, pop_score, hab_score, clim_score, k_cluster, h_cluster) %>%
  mutate(pop_score = round(pop_score, 3),
         hab_score = round(hab_score, 3),
         clim_score = round(clim_score, 3))

sticky_style <- list(backgroundColor = "#f7f7f7")

htmltools::browsable(
  tagList(
    tags$button(
      tagList(fontawesome::fa("download"), "Download as CSV"),
      onclick = "Reactable.downloadDataCSV('download-table', 'Table.csv')"
    ),
    
    # Render table
    reactable::reactable(
      tab,
      bordered = TRUE,
      highlight = TRUE,
      striped = TRUE,
      resizable = TRUE, 
      wrap = TRUE,
      searchable = TRUE,
      style = list(fontSize = "1.25rem"),
      showSortIcon = TRUE,
      height = 800,
      pagination = FALSE,
      elementId = "download-table"
    )
  )
)

```

# References
